---
title: "Coursera Machine Learning Final Project"
author: "Collin S"
date: "August 28, 2016"
output: html_document
---

## Background
This project's goal is to correctly classify the different barbell lifts.*(see source) The particpants did 5 different types of positions. The positions are sitting-down, standing-up, standing, walking, and sitting.  

### load appropriate packages

```{r results = 'hide', message = FALSE, warning = FALSE}
library(caret)
library(rpart)
library(rpart.plot)
library(RColorBrewer)
library(rattle)
library(randomForest)
library(knitr)
```

## Data input and Cleaning

```{r}
# set seed for reproducibility concerns
set.seed(12345)

trainUrl <- "http://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv"
testUrl <- "http://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv"

training <- read.csv(url(trainUrl), na.strings=c('NA','#DIV/0!',''))
testing <- read.csv(url(testUrl), na.strings=c('NA','#DIV/0!',''))
# check if variable names match
# they all match but training has classe variable and testing does not!
names(testing) == names(training)
```

### Split test data into test and training sets for model selection

```{r}
# Partition data into training and test sets 
set.seed(33)
inTrain <- createDataPartition(training$classe, p = .65, list = F)
train1 <- training[inTrain, ]
train2 <- training[-inTrain, ]

# now going to do some data cleaning. This includes excluding unnecessary variables, removing near zero variance data, and data with lots of NAs 
zerovar <- nearZeroVar(train1)
train1 <- train1[, -zerovar]
train2 <- train2[, -zerovar]

mostlyNA <- sapply(train1, function(x) mean(is.na(x))) > 0.80
train1 <- train1[, mostlyNA==F]
train2 <- train2[, mostlyNA==F]

train1 <- train1[, -(1:5)]
train2 <- train2[, -(1:5)]
```

## Fitting a model and cross validation

```{r}
fitControl <- trainControl(method="cv", number=3, verboseIter=F)

fit <- train(classe ~ ., data=train1, method="rf", trControl=fitControl)

# final model
fit$finalModel
```

## Fitting selected model to the entire test set
```{r}
# now go through the same process as before, but include all of the training data. This will give the algorithm more information thus improving the model. 

nzv <- nearZeroVar(training)
training <- training[, -nzv]
testing <- testing[, -nzv]

# remove variables that are almost always NA
mostlyNA <- sapply(training, function(x) mean(is.na(x))) > 0.80
training <- training[, mostlyNA==F]
testing <- testing[, mostlyNA==F]

# Remove variables that are not necessary
training <- training[, -(1:5)]
testing <- testing[, -(1:5)]

# Fit the whole training dataset to the random forest model with a 3 k fold cross validation
fitControl <- trainControl(method="cv", number=3, verboseIter=F)
fit <- train(classe ~ ., data=training, method="rf", trControl=fitControl)

# Test set prediction
predtest <- predict(fit, newdata=testing)
```


## Results
``` {r}
# character vector conversion
predtest <- as.character(predtest)

# function that writes results to a table
pml_write_files <- function(x) {
  n <- length(x)
  for(i in 1:n) {
    filename <- paste0("problem_id_", i, ".csv")
    write.table(x[i], file=filename, quote=F, row.names=F, col.names=F)
  }
}

# prediction file to test with the quiz
pml_write_files(predtest)
```


### Source:
Ugulino, W.; Cardador, D.; Vega, K.; Velloso, E.; Milidiu, R.; Fuks, H. Wearable Computing: Accelerometers' Data Classification of Body Postures and Movements. Proceedings of 21st Brazilian Symposium on Artificial Intelligence. Advances in Artificial Intelligence - SBIA 2012. In: Lecture Notes in Computer Science. , pp. 52-61. Curitiba, PR: Springer Berlin / Heidelberg, 2012. ISBN 978-3-642-34458-9. DOI: 10.1007/978-3-642-34459-6_6.
Read more: http://groupware.les.inf.puc-rio.br/har#ixzz4IjqZqh3W

